Finiteness, Cardinality, and Combinatorics in Homotopy Type Theory

### 2020-01-06 paper.tex
tightened up a little

The term \(x \in \mathit{xs}\) hides a subtle detail of this definition.
Depending on how we choose to represent \(x \in \mathit{xs}\), the resulting
predicate will be equivalent in strength to either manifest Bishop finiteness
(section~\ref{manifest-bishop}) or manifest enumerability
(section~\ref{manifest-enumerability}).
For now, we will define it like so:

\begin{rm-definition}[List Membership] \label{list-membership}
  \begin{equation}
    x \in \mathit{xs} = \sum_{(i : \mathbf{Fin}(\text{length}(\mathit{xs})))} , \mathit{xs} \ind i \equiv x
  \end{equation}
  Here, \(\ind\) is a safe indexing function.
  In other words, to say that some item \(x\) is in some list \(\mathit{xs}\),
  we say that there is an index into \(\mathit{xs}\) which selects an element of
  \(\mathit{xs}\) which is equivalent to \(x\).
\end{rm-definition}
\begin{rm-definition}[\mathbf{Fin}]
  \(\mathbf{Fin}\)(n) is the type of natural numbers smaller than \(n\).
  It is defined inductively, as follows:
  \begin{equation}
    \begin{aligned}
      \mathbf{Fin}(0) &= \bot \\
      \mathbf{Fin}(n + 1) &= \top + \mathbf{Fin}(n)
    \end{aligned}
  \end{equation}
\end{rm-definition}

Clearly, this definition of list membership tells us not just that an item is in
a list, but also its \emph{position} in that list.
That piece of information gives use decidable equality, restricting the
predicate to discrete sets.
In section~\ref{manifest-enumerability}, we will show how to hide the position,
giving us a variant of the predicate which allows for non-set finite types.
\subsection{Containers}
One last part is missing from the finiteness definition above: the definition of
\(\mathbf{List}\).
The standard definition is as follows:
\begin{rm-definition}[Lists as an Inductive Type]
  \begin{equation}
    \begin{aligned}
      \mathbf{List}(A) &= \\
      & \left| \; \text{nil} : \mathbf{List}(A)   \right; \\
      & \left| \; \text{cons} : A \rightarrow \mathbf{List}(A) \rightarrow \mathbf{List}(A) \right;
    \end{aligned}
  \end{equation}
\end{rm-definition}

However, as we will see, it will be occasionally more convenient to work with an
equivalent type defined as a \emph{container}.
\begin{rm-definition}[Container] \label{container-def}
  A container \cite{abbottContainersConstructingStrictly2005} is a dependent
  pair of shapes, and positions indexed by those shapes. 
  \begin{equation}
    \mathbf{Container} = \sum_{(\text{shape} : \text{Type})} , \text{shape} \rightarrow \text{Type}
  \end{equation}
  We ``interpret'' a container into a pair of a specific shape, and a function
  which goes from positions to contents.
  \begin{equation}
    \llbracket S , P \rrbracket = \prod_{(X : \text{Type})} , \sum_{(s : S)} , \left( P(s) \rightarrow X \right)
  \end{equation}
  Membership of a container can be defined like so:
  \begin{equation}
    x \in \mathit{xs} = \text{fiber}(\text{snd}(\mathit{xs}), x)
  \end{equation}
\end{rm-definition}

We often use containers to define types like lists or trees in a more generic
way.
While they tend to be more cumbersome to work with (especially for termination
checking) they have a number of distinct advantages.
In particular, when finiteness predicates are defined in terms of containers,
certain proofs (we will point them out as we come upon them) become trivial.

Happily, we can have our cake and eat it too.
Since we have access to the univalence axiom, we can transport along any
equivalence to a more convenient representation with good justification.

\begin{rm-lemma}
  The two definitions of lists---the traditional inductive type, and the
  container---are equivalent.
  \begin{equation}
    \mathbf{List} \simeq \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket
  \end{equation}
\end{rm-lemma}
\begin{proof}
  \todo[inline]{Agda code for proof?}
\end{proof}

For each finiteness predicate, we have proven it equivalent to its
container-based version.
These proofs are included in the Agda formalisation.
\subsection{Split Surjections}
Another, equivalent way to define ``finiteness'' is via a (split) surjection
from a finite prefix of the natural numbers.
In this section, we will prove that equivalence, formally.

\begin{rm-theorem} \label{split-enum-surj}
  Split enumerability is equivalent to a split surjection from a finite prefix
  of the natural numbers.
  \begin{equation}
    \mathcal{E}!(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{rm-theorem}
\begin{proof}
  For this proof, we will work with the container-based finiteness predicate.
  The proof is surprisingly short: after sufficient inlining, it emerges that
  our goal is simply a reassociation.
  \ExecuteMetaData[agda/Cardinality/Container/Finite.tex]{split-surj}
\end{proof}
\subsection{Decidable Equality}
\begin{rm-lemma} \label{split-enum-discrete}
  Any split enumerable type has decidable equality (is discrete).
\end{rm-lemma}
\begin{proof}
  We use a corollary that if there is a split-surjection from \(A\) to \(B\),
  and \(A\) is discrete, then \(B\) is also discrete.
\end{proof}
\begin{rm-lemma}
  Any split enumerable type is a set.
\end{rm-lemma}
\begin{proof}
  By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, since split
  enumerable types have decidable equality
  (proposition~\ref{split-enum-discrete}), they are sets.
\end{proof}
\subsection{Closure}
In this section we will prove closure under various operations for split
enumerable sets.
We are working towards a topos proof, which requires us to prove closure under
a variety of operations: for now, we only have enough machinery to demonstrate
the semiring operations, and dependent sums.
in order to show closure under exponentials (function arrows), we will need an
equivalence with \(\mathbf{Fin}\), which will be provided in
section~\ref{manifest-bishop}.
\begin{rm-lemma}
  \(\bot\) is split enumerable.
\end{rm-lemma}
\begin{proof}
  The support list for \(\bot\) is the empty list.
  The cover proof is present via the principle of explosion.
\end{proof}
\begin{rm-lemma}
  \(\top\) is split enumerable.
\end{rm-lemma}
\begin{proof}
  The support list for \(\top\) is the singleton list containing \(\text{tt}\).
  The cover proof is a constant function which returns an index into the first
  item in the list.
\end{proof}
Both of those cases were relatively simple.
Next, we will look into how to combine proofs of split enumerability.
\begin{rm-theorem}
  Split-enumerability is closed under \(\sum\).
\end{rm-theorem}
\begin{proof}
  Let \(E_A\) be a proof of split enumerability for some type \(A\), and \(E_U\)
  be a function of the type:
  \begin{equation}
    E_U : \prod_{(x : A)} , \mathcal{E}!(U(x))
  \end{equation}
  In other words, a function which returns a proof of split enumerability for
  each member of the family \(U\).

  To obtain the support list, we concatenate the support lists of all the proofs
  of split-finiteness for \(U\) over the support list of \(E_A\).
  In Agda:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{sigma-enum}
  ``do-notation'' is available to us as we're working in the list monad.

  The cover proof follows in a simple, if tedious, way.
\end{proof}
\begin{rm-lemma}
  Split-enumerability is closed under Cartesian product (non-dependent product).
\end{rm-lemma} 
\begin{proof}
\end{proof}
\begin{rm-lemma}
  Split-enumerability is closed under disjoint union (non-dependent sum) and
  Cartesian product (non-dependent product).
\end{rm-lemma}
\begin{proof}
  Both of these closures can be derived from closure of the \(\sum\) type.
  The non-dependent product is equivalent to a special case of \(\sum\):
  \begin{equation}
    A \times B \simeq \sum_{(x : A)} , B
  \end{equation}
  And disjoint union between two types \(A\) and \(B\) can be represented by the
  following type:
  \begin{equation}
    A + B = \sum_{(x : \mathbf{Bool})} , \text{if} \; x \; \text{then} \; A \; \text{else} \; B
  \end{equation}
  Then, since all of \(\mathbf{Bool}\), \(A\), and \(B\) are split enumerable,
  the type \(A + B\) is split enumerable.
\end{proof}
\section{Manifest Bishop Finiteness} \label{manifest-bishop}
\begin{rm-definition}[Manifest Bishop Finiteness]
  \begin{equation}
    \mathcal{B}(A) = \sum_{(\mathit{xs} : \textbf{List}(A))} , \prod_{(x : A)} , x \inunique xs
  \end{equation}
\end{rm-definition}
The only difference between this predicate and split enumerability is the list
membership term: we use \(\inunique\) here, where \(x \inunique \mathit{xs}\) is
to be read as ``\(x\) occurs exactly once in \(\mathit{xs}\)''.
It is defined in terms of our previous list membership term
(definition~\ref{list-membership}).
\begin{rm-definition}[Unique List Membership]
  We say an item \(x\) is ``uniquely in'' some list \(\mathit{xs}\) if its
  membership in that list is a \emph{contraction}; i.e. its membership proof
  exists, and all such proofs are equal.
  \begin{equation}
    x \inunique \mathit{xs} = \text{isContr}(x \in \mathit{xs})
  \end{equation}
\end{rm-definition}

>>>>

We will at this point take a moment to define some of the types we used to
define \(\mathcal{E}!\).
Lists, and membership thereof, are defined using \emph{containers}.
\begin{rm-definition}[Container] \label{container-def}
  A container \cite{abbottContainersConstructingStrictly2005} is a dependent
  pair of shapes, and positions indexed by those shapes. 
  \begin{equation}
    \mathbf{Container} = \sum_{(\text{shape} : \text{Type})} , \text{shape} \rightarrow \text{Type}
  \end{equation}
  We ``interpret'' a container into a pair of a specific shape, and a function
  which goes from positions to contents.
  \begin{equation}
    \llbracket S , P \rrbracket = \prod_{(X : \text{Type})} , \sum_{(s : S)} , \left( P(s) \rightarrow X \right)
  \end{equation}
  Membership of a container can be defined like so:
  \begin{equation}
    x \in \mathit{xs} = \text{fiber}(\text{snd}(\mathit{xs}), x)
  \end{equation}
\end{rm-definition}
Containers can be used to define a wide variety of functors (streams, trees,
etc.): lists are all that interest us now.
\begin{rm-definition}[\mathbf{Fin}]
  \(\mathbf{Fin}\)(n) is the type of natural numbers smaller than \(n\).
  It is defined inductively, as follows:
  \begin{equation}
    \begin{aligned}
      \mathbf{Fin}(0) &= \bot \\
      \mathbf{Fin}(n + 1) &= \top + \mathbf{Fin}(n)
    \end{aligned}
  \end{equation}
\end{rm-definition}
\begin{rm-definition}[Lists]
  The ``shape'' of lists is \(\mathbb{N}\), indicating the length of the list in
  question.
  \begin{equation}
    \mathbf{List} = \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket
  \end{equation}
\end{rm-definition}

Internally, in our formalisation, we actually use the standard
inductive definition of lists more often (it tends to work better in more
complex algorithms, and functions on it seems to satisfy the termination
checker more readily).
However, since both types are equivalent, univalence allows us to transport to
whichever representation is more convenient in a given situation.
For the higher-level proofs we present here, though, the container-based
definition greatly simplifies certain steps, which is why we have chosen it as
our representation.
\subsection{Split Surjections}
Another, equivalent way to define ``finiteness'' is via a (split) surjection
from a finite prefix of the natural numbers.
In this section, we will prove that equivalence, formally.

\begin{rm-theorem} \label{split-enum-surj}
  Split enumerability is equivalent to a split surjection from a finite prefix
  of the natural numbers.
  \begin{equation}
    \mathcal{E}!(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{rm-theorem}
\begin{proof}
  The proof is surprisingly short: after sufficient inlining, it emerges that
  our goal is simply a reassociation.
  \ExecuteMetaData[agda/Cardinality/Container/Finite.tex]{split-surj}
\end{proof}
\subsection{Decidable Equality}
\begin{rm-lemma} \label{split-enum-discrete}
  Any split enumerable type has decidable equality (is discrete).
\end{rm-lemma}
\begin{proof}
  We use a corollary that if there is a split-surjection from \(A\) to \(B\),
  and \(A\) is discrete, then \(B\) is also discrete.
\end{proof}
\begin{rm-lemma}
  Any split enumerable type is a set.
\end{rm-lemma}
\begin{proof}
  By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, since split
  enumerable types have decidable equality
  (proposition~\ref{split-enum-discrete}), they are sets.
\end{proof}
\subsection{Closure}
In this section we will prove closure under various operations for split
enumerable sets.
We are working towards a topos proof, which requires us to prove closure under
a variety of operations: for now, we only have enough machinery to demonstrate
the semiring operations, and dependent sums.
in order to show closure under exponentials (function arrows), we will need an
equivalence with \(\mathbf{Fin}\), which will be provided in
section~\ref{manifest-bishop}.
\begin{rm-lemma}
  \(\bot\) is split enumerable.
\end{rm-lemma}
\begin{proof}
  The support list for \(\bot\) is the empty list.
  The cover proof is present via the principle of explosion.
\end{proof}
\begin{rm-lemma}
  \(\top\) is split enumerable.
\end{rm-lemma}
\begin{proof}
  The support list for \(\top\) is the singleton list containing \(\text{tt}\).
  The cover proof is a constant function which returns an index into the first
  item in the list.
\end{proof}
Both of those cases were relatively simple.
Next, we will look into how to combine proofs of split enumerability.
\begin{rm-theorem}
  Split-enumerability is closed under \(\sum\).
\end{rm-theorem}
\begin{proof}
  Let \(E_A\) be a proof of split enumerability for some type \(A\), and \(E_U\)
  be a function of the type:
  \begin{equation}
    E_U : \prod_{(x : A)} , \mathcal{E}!(U(x))
  \end{equation}
  In other words, a function which returns a proof of split enumerability for
  each member of the family \(U\).

  To obtain the support list, we concatenate the support lists of all the proofs
  of split-finiteness for \(U\) over the support list of \(E_A\).
  In Agda:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{sigma-enum}
  ``do-notation'' is available to us as we're working in the list monad.
\end{proof}
\begin{rm-lemma}
  Split-enumerability is closed under disjoint union (non-dependent sum) and
  Cartesian product (non-dependent product).
\end{rm-lemma}
\begin{proof}
  Both of these closures can be derived from closure of the \(\sum\) type.
  The non-dependent product is equivalent to a special case of \(\sum\):
  \begin{equation}
    A \times B \simeq \sum_{(x : A)} , B
  \end{equation}
  And disjoint union between two types \(A\) and \(B\) can be represented by the
  following type:
  \begin{equation}
    A + B = \sum_{(x : \mathbf{Bool})} , \text{if} \; x \; \text{then} \; A \; \text{else} \; B
  \end{equation}
  Then, since all of \(\mathbf{Bool}\), \(A\), and \(B\) are split enumerable,
  the type \(A + B\) is split enumerable.
\end{proof}
\section{Manifest Bishop Finiteness} \label{manifest-bishop}
\begin{rm-definition}[Manifest Bishop Finiteness]
  \begin{equation}
    \mathcal{B}(A) = \sum_{(\mathit{xs} : \textbf{List}(A))} , \prod_{(x : A)} , x \inunique xs
  \end{equation}
\end{rm-definition}
The only difference between this predicate and split enumerability is the list
membership term: we use \(\inunique\) here, where \(x \inunique \mathit{xs}\) is
to be read as ``\(x\) occurs exactly once in \(\mathit{xs}\)''.
\begin{rm-definition}[Unique Membership]
  We say an item \(x\) is ``uniquely in'' some container \(\mathit{xs}\) if its
  membership in that list is a \emph{contraction}; i.e. its membership proof
  exists, and all such proofs are equal.
  \begin{equation}
    x \inunique \mathit{xs} = \text{isContr}(x \in \mathit{xs})
  \end{equation}
\end{rm-definition}
